A Constraint Solving Approach to Parikh Images of Regular Languages

### 2023-10-21 calculus.tex
corrected flow
Finally 23 pages

The rule~$\Split{}$ allows us to branch the proof tree by trying to exclude a
contested transition from a potential solution before concluding that it must be
included. Intuitively, this is what guarantees our ability to make forward
progress by eliminating paths through~$\Automaton$.

@@

Starting with~$\mathcal{A}'$, $\Map$ extracts the increments of a transition, $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$. Initially, we let $\Filter$ map to fresh variables to obtain, after some simplifications:
\begin{equation}\label{eq:single-filter}
  \begin{array}{rl@{\qquad}rl@{\qquad}rl}
    \Filter(\FromLabelTo{D}{d/[l_{A+}, n_+]}{F_A})
    &= \TransitionVar_1
    &
    \Filter(\FromLabelTo{F_A}{\Sigma/[ l_{A+} ]}{F_A} )
    &= \TransitionVar_2   
    &
    \Filter(\FromLabelTo{I_A}{c/[l_{A+}, n_+]}{F_A} ) 
    &=  \TransitionVar_3 
    \\
    \Filter(\FromLabelTo{I_A}{d/[l_{A+}, n_+]}{D}) 
    &= \TransitionVar_4
    &
    \Filter(\FromLabelTo{I_A}{\Sigma/[ l_{A+}, i_+]}{I_A}) 
    &= \TransitionVar_5  \\[3ex]
%
  \multicolumn{6}{c}{
    \FlowEq(\mathcal{A}', \Filter)  = \begin{aligned}
                       1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_4 + \TransitionVar_3\\
                    \TransitionVar_4 &= \TransitionVar_1 \\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &= \cancel{\TransitionVar_2} + 1  \\
                    \end{aligned}}
  \end{array}
\end{equation}

>>>>

The rule~$\Split{}$ allows us to branch the proof tree by trying to exclude a
transition from a potential solution before concluding that it must be included.
Intuitively, this is what guarantees our ability to make forward progress by
eliminating paths through~$\Automaton$.

@@

Starting with~$\mathcal{A}'$, $\Map$ extracts the increments of a transition, $h(\FromLabelTo{q_1}{c / i}{q_2}) = \Vec{v}$ for the increment
vector $\Vec{v}$, concretely $h(\FromLabelTo{I_A}{\Sigma / \left[ l_{A+}, i_+
\right]}{I_A}) = \left[ l_{A+}, i_+ \right]$. We use the same compact notation
here as in \cref{sec:intuition} to represent what is essentially a sparse vector
of $1$ and $0$ coefficients, i.e. $\left[ l_{A+}, i_+ \right] = \left[1, 1,
0\right]$.  The reader is
advised to review \cref{fig:aut_a} from
\cref{sec:intuition} while going through this example.

Initially, we let $\Filter$ map to fresh variables to obtain, after some simplifications:
\begin{equation}\label{eq:single-filter}
  \begin{array}{rl@{\qquad}rl@{\qquad}rl}
    \Filter(\FromLabelTo{D}{d/[l_{A+}, n_+]}{F_A})
    &= \TransitionVar_1
    &
    \Filter(\FromLabelTo{F_A}{\Sigma/[ l_{A+} ]}{F_A} )
    &= \TransitionVar_2   
    &
    \Filter(\FromLabelTo{I_A}{c/[l_{A+}, n_+]}{F_A} ) 
    &=  \TransitionVar_3 
    \\
    \Filter(\FromLabelTo{I_A}{d/[l_{A+}, n_+]}{D}) 
    &= \TransitionVar_4
    &
    \Filter(\FromLabelTo{I_A}{\Sigma/[ l_{A+}, i_+]}{I_A}) 
    &= \TransitionVar_5  \\[3ex]
%
  \multicolumn{6}{c}{
    \FlowEq(\mathcal{A}', \Filter)  =\quad \begin{aligned}
                       1 + \cancel{\TransitionVar_5} &= \cancel{\TransitionVar_5} + \TransitionVar_4 + \TransitionVar_3 \wedge\mbox{}\\
                    \TransitionVar_4 &= \TransitionVar_1 \wedge\mbox{}\\
                    \TransitionVar_3 + \TransitionVar_1 + \cancel{\TransitionVar_2} &= \cancel{\TransitionVar_2} + 1  \\
                    \end{aligned}}
  \end{array}
\end{equation}

### 2023-10-21 implementation.tex
Remove dangling references, add fig 6 description
Finally 23 pages

\Catra{} uses symbolic labels for automata. A symbolic label is defined as a
finite range of Unicode code points. This allows succinct representation of many
regular expression patterns such as \lstinline{(a-z).*}, which would have
otherwise required $27$~transitions. The transition for the range would be
written \lstinline{a -> b [97, 122]}.

In satisfaction mode, supported by all backends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting \Sat{} with register
assignments or \Unsat{} much like traditional~SAT- or SMT solvers would.
Additionally, baseline and \Calculus{} also support generating the Presburger
formula describing the constraints of the input file, i.e., computing
a closed-from representation of the complete Parikh image.
% Baseline uses standard
%quantifier elimination, and \Calculus{} uses the method described in
%\cref{sec:finding-the-image}.

Since \Princess{} does not support the multiple-arity predicates used
in \Calculus{}, we have implemented variable-length arguments using
additional helper predicates. These are $\Unused{}(\Automaton)$, which
marks an automaton as unused in any product, and
$\TransitionMask{}(\Automaton, \Transition, \Filter(\Transition))$
which associates a transition $\Transition$ and automaton $\Automaton$
with its corresponding transition
variable~$\Filter(\Transition)$. Additionally, we associate each
predicate with an instance variable to tie helper predicates to the
$\Image$ predicate they originated from.

@@

  $q \gets \NewPriorityQueue{}$

@@

\Catra{} enables clause learning by default when using our backend, as it has
been experimentally shown to increase the performance in aggregate (though not
strictly). We do not currently implement all the proposed features of
\cref{sec:ext:backjumping}, but we do implement forward-reachability cut
learning. No sophisticated clause learning for products has been implemented.

>>>>

\Catra{} uses symbolic labels for automata. A symbolic label is defined as a
finite range of Unicode code points. This allows representing regular expression
patterns like \lstinline{(a-z)} as \lstinline{a -> b [a, z]} when it would have
otherwise required $27$~non-symbolic transitions.

In satisfaction mode, supported by all backends, \Catra{} tries to satisfy the
constraints expressed by the input file, reporting \Sat{} with register
assignments or \Unsat{}. Additionally, baseline and \Calculus{} also support
generating the Presburger formula describing the constraints of the input file,
i.e., computing a closed-form representation of the complete Parikh image.
% Baseline uses standard
%quantifier elimination, and \Calculus{} uses the method described in
%\cref{sec:finding-the-image}.

% Since \Princess{} does not support the multiple-arity predicates used
% in \Calculus{}, we have implemented variable-length arguments using
% additional helper predicates. These are $\Unused{}(\Automaton)$, which
% marks an automaton as unused in any product, and
% $\TransitionMask{}(\Automaton, \Transition, \Filter(\Transition))$
% which associates a transition $\Transition$ and automaton $\Automaton$
% with its corresponding transition
% variable~$\Filter(\Transition)$. Additionally, we associate each
% predicate with an instance variable to tie helper predicates to the
% $\Image$ predicate they originated from.

@@

  $q \gets \NewPriorityQueue{$\mathcal{A}_i, \ldots, \mathcal{A}_n$}$

@@

\Catra{} enables clause learning by default when using our backend, as it has
been experimentally shown to increase the performance in aggregate (though not
strictly). We currently only implement minimal clause learning based on
forward-reachability cuts. No sophisticated clause learning for products has
been implemented.
